Enabling inference for context-dependent models of mutation by bounding the propagation of dependency

### 2022-01-18 writeups/context-dependence.tex
Changing the title
authors and funding
Adding reviews and suggestions

\title{Inference using context-dependent models of mutation via bounds on propagation of dependency}
\author[1]{Peter L Ralph}
\author[2,3,4,5]{Frederick A Matsen IV}
\author[1,*]{Peter L Ralph}

@@

FAM's work supported by NIH grants R01 AI146028 and U01 AI150747.
Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.

@@

Fix one tip of the tree, $a$, to be the taxa where we count ``long'' sequences, let $\rho$ be the root, and let $b_1, \ldots, b_n$ be the remaining tips,
where we observe ``short'' sequences.
Let $v_0=\rho, v_1, \ldots, v_\ell = a$ be the path from the root down to $a$,
let $u_1, u_2, \ldots, u_m$ be the remaining internal nodes,
and for each $v_i$ let $u(v_i)$ be the offspring of $v_i$ that is not one of the other $v$.
For each node in the tree,
we will compute a matrix of probabilities:
for nodes not on the path from $a$ to the root,
we compute the chance that that each ``long'' sequence observed at that node
is matched by each possible combination of ``short'' sequences on the tips of the tree below it.
For nodes on the path from $a$ to the root, we compute the probability of observing
each ``long'' sequence at that node, along with the combinations of ``short'' sequences
on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, denoted $M(w)$,
where $c$ is the number of tips for which $w$ lies on the path from $a$ to the tip.
Rows of $M(u_j)$ sum to 1,
while each entire matrix $M(v_i)$ sums to 1;
the difference being that $M(v_i)$ sums over the prior at the root,
and $M(u_j)$ takes the value at $u_j$ as given.
Below, we index columns of $M(u)$ by lists of short patterns,
and leave the specific ordering to implementation.
(For instance, if $c=2$ then a generic entry is $M(u)_{x,(y,z)}$,
where the long pattern $x$ indexes the rows and the pair of short patterns $(y,z)$ indexes the columns;
in the software all matrices retain column names making this unambiguous.)

>>>>

\title{Enabling inference for context-dependent models of mutation by bounding the propagation of dependency}
\author[2,3,4,5]{Frederick A Matsen IV}
\author[1,*]{Peter L Ralph}

@@

FAM's work supported by NIH grants R01 AI146028 and U01 AI150747.
Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.
PLR's work was supported by NIH award R01HG010774

@@

Fix one tip of the tree, $a$, to be the taxa where we count ``long'' sequences, let $\rho$ be the root, and let $b_1, \ldots, b_n$ be the remaining tips,
where we observe ``short'' sequences.
Let $v_0=\rho, v_1, \ldots, v_\ell = a$ be the path from the root down to $a$,
let $u_1, u_2, \ldots, u_m$ be the remaining internal nodes,
and for each $v_i$ let $u(v_i)$ be the offspring of $v_i$ that is not one of the other $v$.
For each node in the tree,
we will compute a matrix of probabilities:
for nodes not on the path from $a$ to the root,
we compute the chance that each ``long'' sequence observed at that node
is matched by each possible combination of ``short'' sequences on the tips of the tree below it.
For nodes on the path from $a$ to the root, we compute the probability of observing
each ``long'' sequence at that node, along with the combinations of ``short'' sequences
on all tips further away from $a$ (i.e., below that node if it was rooted at $a$).
For a node $w$ this will be stored as a $|\calS|^{n} \times |\calS|^{cm}$ matrix, denoted $M(w)$,
where $c$ is the number of tips for which $w$ lies on the path from $a$ to the tip.
Rows of $M(u_j)$ sum to 1,
while each entire matrix $M(v_i)$ sums to 1;
the difference being that $M(v_i)$ sums over the prior at the root,
and $M(u_j)$ takes the value at $u_j$ as given.
Below, we index columns of $M(u)$ by lists of short patterns,
and leave the specific ordering to implementation.
(For instance, if $c=2$ then a generic entry is $M(u)_{x,(y,z)}$,
where the long pattern $x$ indexes the rows and the pair of short patterns $(y,z)$ indexes the columns;
in the software all matrices retain column names making this unambiguous.)
